Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported b

### 2021-09-14 paper.Rmd
various mostly formal additions

---
title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
author: |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}

@@

<!-- % emotion in political science -->
Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions. 
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization]. 
Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude]. 
While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].

@@

@@

A study has shown that moral shock works to moralize attitudes towards GM food.
However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

@@

To investigate the direct effects of misinfomation and corrective information conditional on

@@

## Treatment Material

## Outcome measures

@@

@@

Table \@ref(tab:descriptives) shows

```{r descriptives, warning=FALSE, echo=FALSE, fig.pos="H"}
# only control group

descriptives_control <- data %>% 
  filter(exp_group == "control") %>% 
  select("Anger" = "anger",
         "Anxiety" = "anxiety",
         "Disgust" = "disgust",
         "Sadness" = "sadness",
         "Belief GMO safety" = "knowledge_safe", 
         "Policy restrictions" = "policy_opinion",
         "Moral conviction" = "moral_conviction")

knitr::kable(datasummary_skim(descriptives_control, histogram = FALSE, output = "data.frame"), 
             caption = "Descriptive statistics")

@@

@@

@@

>>>>

---
title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
author: |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

<!-- % emotion in political science -->
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization]. 
Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude].

@@

Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.

@@

A study has shown that moral shock works to moralize attitudes towards GM food.

While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].

However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

@@

To investigate the direct effects of misinformation and corrective information conditional on

@@

## Treatment Material {-}

The experimental treatment consists of a combination of two types of information. One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats. The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional. The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed. The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128]. If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.

The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon. This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information @[lewandowsky2020debunking]. As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format. By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food. This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.

Assignment is done by simple random assignment into equal sized groups. Table gives an overview over the nine resulting experimental groups. The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned). For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' @[kuklinski2000misinformation, p.805].

## Outcome measures {-}

As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct".

 To provide a broad measure of policy support, I use a measure by @[diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".

I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" @[skitka2020psychology]. This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them. Both items will be combined into a simple additive index.

I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food. Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds @[skitka2018attitude]. The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".

## Ethics {-}

An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
Because all shown information has in a similar format been in the media, there is no deception taking place.
The experimental treatment intends to evoke emotions, but I consider them within reason. There are two main reasons for that. First, the material including the used imagery has been used in a previously published study @[clifford2019emotional]. Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents. Because this approval has been granted, this study did not require further approval by an ethics committee.

@@

This paper is written in RMarkdown based on a template available here: https://osf.io/q395s

@@

Table \@ref(tab:descriptives)

```{r descriptives, echo=FALSE, message=FALSE, warning=FALSE}
library(modelsummary)
# only control group

data %>% 
  filter(exp_group == "control") %>% 
  select("Anger" = "anger",
         "Anxiety" = "anxiety",
         "Disgust" = "disgust",
         "Sadness" = "sadness",
         "Belief GMO safety" = "knowledge_safe", 
         "Policy restrictions" = "policy_opinion",
         "Moral conviction" = "moral_conviction") %>% 
  datasummary_skim(histogram =FALSE,
                   output = "default",
                   title = "Descriptive statistics in the control group")

@@

\newpage
\pagenumbering{roman}

@@

## Information effect

@@

## Regression tables

```{r reg-output-emotion, message=FALSE, warning=FALSE, echo = FALSE}

library(modelsummary)

cm <- c('(Intercept)'    = 'Control',
        'exp_grouplow-misinfo'    = 'Low emotion misinf.',
        'exp_grouphigh-misinfo' = 'High emotion misinf.',
        'exp_grouplow-misinfo-neutral-correction' = 'Neutral correction to low emotion misinf.',
        'exp_grouphigh-misinfo-neutral-correction' = 'Neutral correction to high emotion misinf.',
        'exp_grouplow-misinfo-counter-frame' = 'Counterframe low emotion misinf.',
        'exp_grouphigh-misinfo-counter-frame' = 'Counterframe to high emotion misinf.',
        'exp_groupneutral-correction'    = 'Neutral correction w/o misinf.',
        'exp_groupcounter-frame'    = 'Counterframe w/o misinf.')

models_emotions <- list("Anger" = lm_emotion_anger,
                        "Anxiety" =  lm_emotion_anxiety,
                        "Disgust" = lm_emotion_disgust,
                        "Sadness" = lm_emotion_sadness)

modelsummary(models_emotions,
             coef_map = cm,
             title = 'Regression table for emotions.', 
             stars = c('*' = .05, '**' = .01, '***' = 0.01),
             gof_omit = 'DF|Deviance|F|Log|AIC|BIC',
             output = 'markdown')

```
\newpage

```{r reg-output-belief-opinion, message=FALSE, warning=FALSE, echo = FALSE}

models <- list("Belief GMO safety" = lm_knowledge_safe,
               "Policy restrictions" =  lm_policy_opinion,
               "Moral conviction" = lm_moral_conviction)

modelsummary(models,
             coef_map = cm,
             title = 'Regression table for belief and policy opinion.', 
             stars = c('*' = .05, '**' = .01, '***' = 0.01),
             gof_omit = 'DF|Deviance|F|Log|AIC|BIC',
             output = 'markdown')

## Treatment material

```{r screenshot-low-misinfo, fig.align = 'center', out.width = "100%", fig.cap = "Low emotion misinformation", echo = FALSE}
knitr::include_graphics("./figures/misinfo_low_emotion.png")
```

```{r screenshot-high-misinfo, fig.align = 'center', out.width = "100%", fig.cap = "High emotion misinformation", echo = FALSE}
knitr::include_graphics("./figures/misinfo_high_emotion.png")
```

```{r screenshot-neutral-correction, fig.align = 'center', out.width = "100%", fig.cap = "Neutral correction", echo = FALSE}
knitr::include_graphics("./figures/neutral_correction.png")
```

```{r screenshot-counterframe, fig.align = 'center', out.width = "100%", fig.cap = "Counterframe correction", echo = FALSE}
knitr::include_graphics("./figures/counter_frame.png")
```
